---
title: "SQL Server 2016 R Services NYC Taxi Demo"
output: html_document
---

# Abstract 

This is an updated version of the original SQL Server 2016 R Services Walkthrough using 
SQL Server 2016 R Services. The original version of the document can be found 
[here](https://msdn.microsoft.com/en-us/library/mt612857.aspx).

It has been updated to be shippable as a self-contained R Tools for Visual Studio project.
The code for this project can always be found at this [Github repo](https://github.com/jflam/IgniteDemo).

All of the sample code has been tested and verified to work with the following versions
of the software:

* SQL Server 2016 RTM
* R Services for SQL Server 2016 RTM, which uses R 3.2.5 
* R Tools for Visual Studio 2015 v0.5

# Introduction

In this walkthrough, you'll develop an end-to-end solution for predictive modeling using 
R Services (In-database).

This walkthrough is based on a well-known public data set, the New York City taxi dataset. 
You will use a combination of R code, SQL Server data, and custom SQL functions to build 
a classification model that indicates the probability that the driver will get a tip on a 
particular taxi trip. You'll also deploy your R model to SQL Server and use server data to 
generate scores based on the model.

This example is chosen because it can easily be extended to all kinds of real-life problems, 
such as predicting customer responses to sales campaigns, or predicting spending by visitors 
to events.

Because the model can be invoked from a stored procedure, you can also easily embed it in 
an application.

# Pre-Requisites

This tutorial assumes that you already have SQL Server 2016 with R Services installed on
your computer. You can download the free Developer Edition of SQL Server 2016, which
also contains the R Services functionality from 
[this link](https://www.microsoft.com/en-us/cloud-platform/sql-server-editions-developers).

Detailed instructions on how to setup R Services within an existing SQL Server 2016 
installation can be found in [this article](https://msdn.microsoft.com/en-us/library/mt637136.aspx).

# Importing the data into the database

TODO: it is probably a good idea to make the SQL Server backup file for a database 
available, much the same way that the AdventureWorks database is made available for 
download.


# Validating that the database is imported correctly

The code in the blocks below should all print out TRUE if you have 
imported the database correctly. Also, it validates that both the
RevoScaleR library, which is part of Microsoft R Client, and the 
RODBC library are installed and functioning correctly.

```{r}

# Validate that the connection string to the database is working,
# and that both the RevoScaleR and RODBC libraries are functioning.

source("dependencies.R")
source("settings.R")

library(RUnit)
library(RevoScaleR)

data <- RxSqlServerData(
    sqlQuery = "select top 10 * from nyctaxi_sample",
    connectionString = dbConnection)

localFile <- file.path(tempdir(), "sql_data.xdf")
rxImport(data, 
    localFile, 
    overwrite = TRUE, 
	reportProgress = 0)
df <- rxDataStep(localFile, reportProgress = 0)
checkEquals(10, length(df[, 1]))

library(RODBC)

conn <- odbcDriverConnect(connection = dbConnection)
df <- sqlQuery(conn, query = "select count(*) from nyctaxi_sample")
checkEquals(1703957, df[1,1])
```

# Exploring the data

This code is used to run code in the R Server context within SQL Server to summarize 
the data for the 1% down-sampled dataset using the RxSummaryInfo function. The key
thing here is that the execution all happens on the SQL server, and the data is 
never transmitted to the client.

```{r}

# Establish a connection to the SQL Server R Services compute context

sqlShareDir <- paste("C:\\AllShare\\", Sys.getenv("USERNAME"), sep = "")
sqlWait <- TRUE
sqlConsoleOutput <- FALSE

cc <- RxInSqlServer(
    connectionString = dbConnection, 
	shareDir = sqlShareDir,
    wait = sqlWait, 
    consoleOutput = sqlConsoleOutput)

rxSetComputeContext(cc)

# Execute the following SQL query

sampleDataQuery <- "SELECT 
	tipped, fare_amount, passenger_count, trip_time_in_secs,
	trip_distance, pickup_datetime, dropoff_datetime, 
	CAST(pickup_longitude as float) AS pickup_longitude, 
	CAST(pickup_latitude as float) AS pickup_latitude, 
	CAST(dropoff_longitude as float) AS dropoff_longitude, 
	CAST(dropoff_latitude as float) AS dropoff_latitude 
FROM nyctaxi_sample"

inDataSource <- RxSqlServerData(
    sqlQuery = sampleDataQuery, 
	connectionString = dbConnection,
    colClasses = c(
        pickup_longitude = "numeric", 
		pickup_latitude = "numeric",
        dropoff_longitude = "numeric", 
        dropoff_latitude = "numeric"),
    rowsPerRead = 500)

# Compute summary statistics for fares based on passenger count
# in each ride. This computation takes place on the server and can
# take some time to run.

results <- rxSummary( ~ fare_amount:F(passenger_count, 1, 6), data = inDataSource)

library(DT)
datatable(results$categorical[[1]])
```